Survival Compass\thanks{Code and data are available at: https://github.com/LexiKnight/Lung\_Cancer/tree/main}

### 2024-04-15 paper/paper.qmd
Fixed modeling section (formating) and working on data section (started writing)

---
title: "Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis"
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long
abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Linear Regression Model 
# interested in predicting days_to_death, a continuous outcome variable based on
# multiple predictor variables  (pathogenic_stage, lymph_node_involvement, 
# presence_of_distant_metastasis, tumor_size, treatment_decision, treatment_type)
# Linear regression model are suitable for predicting continuous outcomes
# quantifying relationship between predictor variables and outcome variable.

# Fit a linear regression model to predict continuous outcome variable 
# 'days_to_death'

## Model set-up
In this study, we aim to investigate the survival outcomes of lung cancer patients post-diagnosis. Let $y_i$ denote the number of days from diagnosis to death for patient $i$. We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.

We model the survival time $y_i$ using a linear regression framework where $mi_i$ is the expected survival time for patient $i$, and $σ$ represents the standard deviation of the survival times. The linear predictor $mu_i$ is specified as:

\begin{align*}
y_i | \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
\end{align*}

Here, \alpha represents the intercept term, \beta coefficients denote the effect of respective predictors, and subscripts $i$ denote individual patients.

Priors are specified for the model parameters as follows:

\begin{align*}
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

TEMPLATE:
Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align}

\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

@@

The linear regression model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.

>>>>

---
title: "Survival Compass"
subtitle: "Statistical Insights into Lung Cancer Patients Journey Post Diagnosis"
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathogenic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between pathogenic stage, seeking treatment and survival outcomes. Notably, patients at advanced stages with metastases in distant sites beyond the lung, extensive lymph node involvement and tumors with extensive growth, invading nearby structures demonstrate lower survival rates. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

## Model set-up

In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework. We consider several predictors including pathogenic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.

### Model Specifications ###

We employ a linear regression model to predict the number of days from diagnosis to death for each lung cancer patient. The model is defined as follows:

\[
y_i \mid \mu_i, \sigma \sim \text{Normal}(\mu_i, \sigma)
\]

where:
\begin{itemize}
  \item $y_i$ represents the number of days from diagnosis to death for patient $i$.
  \item $mu_i$ denotes the expected number of days to death for patient $i$.
  \item $σ$ represents the standard deviation of the survival times.
\end{itemize}

The linear predictor $mu_i$ is specified as:

\begin{align*}
y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{\text{pathologic\_stage}} \times \text{pathologic\_stage}_i \\
&\quad + \beta_{\text{lymph\_node}} \times \text{lymph\_node\_involvement}_i \\
&\quad + \beta_{\text{metastasis}} \times \text{presence\_of\_distant\_metastasis}_i \\
&\quad + \beta_{\text{tumor\_size}} \times \text{tumor\_size}_i \\
&\quad + \beta_{\text{treatment\_type}} \times \text{treatment\_type}_i
\end{align*}

where: 
\begin{itemize}
  \item \alpha represents the intercept term, capturing the baseline number of days to death.
  \item \beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} are the coefficients associated with each predictor variable.
\end{itemize}

@@

Linear regression models are most appropritate in predicting continuous outcomes. As survival time is continuous, this model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.

#### Response Variable ####

Out variable of interest is survival time in lung cancer patient afte they have been diagnosed

We model the survival time ($y_i$) as a continuous variable, reflecting the duration from diagnosis to death for each patient. This continuous characterization is appropriate for capturing the temporal aspect of survival outcomes in medical contexts.

#### Input Variables ####

We consider several clinically relevant predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type. These variables are chosen based on their established associations with lung cancer prognosis, encompassing key aspects of disease severity and treatment strategies.

#### Model Structure #####

The linear regression model relates the expected survival time ($μ_i$) to the linear combination of predictor variables, allowing us to quantify the impact of each predictor on the expected duration of survival. This framework facilitates interpretation of the associations between clinical variables and survival outcomes, providing valuable insights for patient prognosis.

#### Parameter Estimation ####

@@

We run the model in R [@citeR] estimating the model coefficients ($/alpha$ and $/beta$) using Bayesian inference via the 'stan_glm()' function from the @rstanarm package. This approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain posterior distributions for the model parameters, enabling robust estimation of parameter uncertainties and inference on the effects of predictor variables.

### 2024-04-15 paper/paper.tex
Fixed modeling section (formating) and working on data section (started writing)

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Survival Compass: Statistical Insights into Lung Cancer Patients' Journey Post-Diagnosis},
  pdfauthor={Lexi Knight},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Survival Compass: Statistical Insights into Lung Cancer Patients'
Journey Post-Diagnosis\thanks{Code and data are available at:
https://github.com/LexiKnight/Lung\_Cancer/tree/main}}
\author{Lexi Knight}
\date{April 4, 2024}

\begin{document}
\maketitle
\begin{abstract}
This study investigates the impact of pathogenic stage and treatment
modalities on lung cancer survival post-diagnosis. Analysis of patient
data reveals significant correlation between pathogenic stage, seeking
treatment and survival outcomes. Notably, patients at advanced stages
with metastases in distant sites beyond the lung, extensive lymph node
involvement and tumors with extensive growth, invading nearby structures
demonstrate lower survival rates. These findings underscore the critical
importance of early detection, tailored treatment strategies and ongoing
research efforts to enhance lung cancer survival rates globally.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{linear-regression-model}{%
\subsection{Linear Regression Model}\label{linear-regression-model}}

\hypertarget{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}{%
\section{interested in predicting days\_to\_death, a continuous outcome
variable based
on}\label{interested-in-predicting-days_to_death-a-continuous-outcome-variable-based-on}}

\hypertarget{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}{%
\section{multiple predictor variables (pathogenic\_stage,
lymph\_node\_involvement,}\label{multiple-predictor-variables-pathogenic_stage-lymph_node_involvement}}

\hypertarget{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}{%
\section{presence\_of\_distant\_metastasis, tumor\_size,
treatment\_decision,
treatment\_type)}\label{presence_of_distant_metastasis-tumor_size-treatment_decision-treatment_type}}

\hypertarget{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}{%
\section{Linear regression model are suitable for predicting continuous
outcomes}\label{linear-regression-model-are-suitable-for-predicting-continuous-outcomes}}

\hypertarget{quantifying-relationship-between-predictor-variables-and-outcome-variable.}{%
\section{quantifying relationship between predictor variables and
outcome
variable.}\label{quantifying-relationship-between-predictor-variables-and-outcome-variable.}}

\hypertarget{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}{%
\section{Fit a linear regression model to predict continuous outcome
variable}\label{fit-a-linear-regression-model-to-predict-continuous-outcome-variable}}

\hypertarget{days_to_death}{%
\section{`days\_to\_death'}\label{days_to_death}}

@@

In this study, we aim to investigate the survival outcomes of lung
cancer patients post-diagnosis. Let \(y_i\) denote the number of days
from diagnosis to death for patient \(i\). We consider several
predictors including pathogenic stage, lymph node involvement, presence
of distant metastasis, tumor size, and treatment type.

We model the survival time \(y_i\) using a linear regression framework:

y\_i\textbar{}\mu\_i, \sigma \&\sim \mbox{Normal}(\mu\_i, \sigma)
\textbackslash{}

where \(mi_i\) is the expected survival time for patient \(i\), and
\(σ\) represents the standard deviation of the survival times. The
linear predictor \(mu_i\) is specified as:

\mu\emph{i = \alpha + \beta}\{\text{pathologic\_stage}\}
\times \text{pathologic\_stage}\emph{i + \beta}\{\text{lymph\_node}\}
\times \text{lymph\_node\_involvement}\emph{i +
\beta}\{\text{metastasis}\}
\times \text{presence\_of\_distant\_metastasis}\emph{i +
\beta}\{\text{tumor\_size}\} \times \text{tumor\_size}\emph{i +
\beta}\{\text{treatment\_type}\} \times \text{treatment\_type}\_i

Here, \alpha represents the intercept term, \beta coefficients denote
the effect of respective predictors, and subscripts \(i\) denote
individual patients.

Priors are specified for the model parameters as follows:

\begin{align*}
\alpha &\sim \text{Normal}(0, 2.5) \\
\beta_{\text{pathologic_stage}}, \beta_{\text{lymph_node}}, \beta_{\text{metastasis}}, \beta_{\text{tumor_size}}, \beta_{\text{treatment_type}} &\sim \text{Normal}(0, 2.5) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}

TEMPLATE: Define \(y_i\) as the number of seconds that the plane
remained aloft. Then \(\beta_i\) is the wing width and \(\gamma_i\) is
the wing length, both measured in millimeters.

\begin{align}

\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R (R Core Team 2023) using the \texttt{rstanarm}
package of Goodrich et al. (2022). We use the default priors from
\texttt{rstanarm}.

@@

@@

The linear regression model allows us to quantify the relationships
between these predictors and survival outcomes, providing valuable
insights into the factors influencing the prognosis of lung cancer
patients.

>>>>

\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Survival Compass},
  pdfauthor={Lexi Knight},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Survival Compass\thanks{Code and data are available at:
https://github.com/LexiKnight/Lung\_Cancer/tree/main}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Statistical Insights into Lung Cancer Patients Journey Post
Diagnosis}
\author{Lexi Knight}
\date{April 5, 2024}

\begin{document}
\maketitle
\begin{abstract}
This study investigates the impact of pathogenic stage and treatment
modalities on lung cancer survival post-diagnosis. Analysis of patient
data reveals significant correlation between pathogenic stage, seeking
treatment and survival outcomes. Notably, patients at advanced stages
with metastases in distant sites beyond the lung, extensive lymph node
involvement and tumors with extensive growth, invading nearby structures
demonstrate lower survival rates. These findings underscore the critical
importance of early detection, tailored treatment strategies and ongoing
research efforts to enhance lung cancer survival rates globally.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

@@

In this section, we aim to predict the survival outcomes of lung cancer
patients post-diagnosis with a linear regression model framework. We
consider several predictors including pathogenic stage, lymph node
involvement, presence of distant metastasis, tumor size, and treatment
type.We specify the model and subsequently justify its appropriateness
for our analysis.

\hypertarget{model-specifications}{%
\subsubsection{Model Specifications}\label{model-specifications}}

We employ a linear regression model to predict the number of days from
diagnosis to death for each lung cancer patient. The model is defined as
follows:

\(y_i\) \textbar{} \mu\_i, \sigma \&\sim \text{Normal}(\mu\_i, \sigma)

where: * \(y_i\) represents the number of days from diagnosis to death
for patient \(i\). * \(mu_i\) denotes the expected number of days to
death for patient \(i\). * \(σ\) represesents the standard deviation of
the survival times.

The linear predictor \(mu_i\) is specified as:

\mu\emph{i = \alpha + \beta}\{\text{pathologic_stage}\}
\times \text{pathologic_stage}\emph{i + \beta}\{\text{lymph_node}\}
\times \text{lymph_node_involvement}\emph{i +
\beta}\{\text{metastasis}\}
\times \text{presence_of_distant_metastasis}\emph{i +
\beta}\{\text{tumor_size}\} \times \text{tumor_size}\emph{i +
\beta}\{\text{treatment_type}\} \times \text{treatment_type}\_i

where:

\begin{itemize}
\tightlist
\item
  \alpha represents the intercept term, capturing the baseline number of
  days to death.
\item
  \beta\emph{\{\text{pathologic_stage}\}, \beta}\{\text{lymph_node}\},
  \beta\emph{\{\text{metastasis}\}, \beta}\{\text{tumor_size}\},
  \beta\_\{\text{treatment_type}\} are the coefficients associated with
  each predictor variable.
\end{itemize}

@@

Linear regression models are most appropritate in predicting continuous
outcomes. As survival time is continuous, this model allows us to
quantify the relationships between these predictors and survival
outcomes, providing valuable insights into the factors influencing the
prognosis of lung cancer patients.

\hypertarget{response-variable}{%
\paragraph{Response Variable}\label{response-variable}}

Out variable of interest is survival time in lung cancer patient afte
they have been diagnosed

We model the survival time (\(y_i\)) as a continuous variable,
reflecting the duration from diagnosis to death for each patient. This
continuous characterization is appropriate for capturing the temporal
aspect of survival outcomes in medical contexts.

\hypertarget{input-variables}{%
\paragraph{Input Variables}\label{input-variables}}

We consider several clinically relevant predictors including pathologic
stage, lymph node involvement, presence of distant metastasis, tumor
size, and treatment type. These variables are chosen based on their
established associations with lung cancer prognosis, encompassing key
aspects of disease severity and treatment strategies.

\hypertarget{model-structure}{%
\paragraph{Model Structure}\label{model-structure}}

The linear regression model relates the expected survival time (\(μ_i\))
to the linear combination of predictor variables, allowing us to
quantify the impact of each predictor on the expected duration of
survival. This framework facilitates interpretation of the associations
between clinical variables and survival outcomes, providing valuable
insights for patient prognosis.

\hypertarget{parameter-estimation}{%
\paragraph{Parameter Estimation}\label{parameter-estimation}}

@@

We run the model in R (R Core Team 2023) estimating the model
coefficients (\(/alpha\) and \(/beta\)) using Bayesian inference via the
`stan\_glm()' function from the Goodrich et al. (2022) package. This
approach leverages Markov Chain Monte Carlo (MCMC) algorithms to obtain
posterior distributions for the model parameters, enabling robust
estimation of parameter uncertainties and inference on the effects of
predictor variables.
